---
title: "PADIS vignette"
author: "Kai T. Horstmann"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This package was written for Lara Kammrath from Wake Forest University and her research. It contains some general functions that may apply to multiple, more general problems (section 1) and some very specific functions (section 2) that are tailored to Lara's specific tasks.

It relies on several different packages, which are installed automatically once this package is installed

- tidyr
- data.table
- stringr
- xlsx
- purrr
- psych


## How to install this package:

You can run the following code to install this package from GitHub:


```{r install paids, eval = FALSE, echo = TRUE}


install.packages("devtools")
devtools::install_github("kthorstmann/padis")
library(padis)


```


# Section 1 - General functions

### Example data frame

Th package contains a fake example data frame, `wide_example_data`, which is used to illustrate the functions.

```{r}

library(padis)
data <- wide_example_data
head(data)

```

## Data selection and transformation

### select_vars

This function can be used to select specific variables from a data set. It is primarily build on `grep` or `dplyr::select`.

```{r select_vars}

subset <- select_vars(data, id = "id", prefix="var_s")
head(subset, 3)

subset <- select_vars(data, id = c("id", "id2"), prefix="var_.+3$")
head(subset, 3)

```

To get more information on the use of regular expressions, see `?base::regex`.

### gather_one

```{r gather}

gather_df <- gather_one(data=data, key_vars="id", varying = c("var_p_1", "var_p_2"), new_name="var")
head(gather_df)

```

### swap_char

This function is used to change variable names since the `gather` function (see below, based on `tidyr::gather`) relies on a specific strucutre of item names. It is primarily used in some functions specific to the problem at hand (section 2).

```{r swap_char}

swap_char("this.string", ".", "_")

```


### gather_multiple

Similar to `gather_one`, `gather_multiple` turns several variables from a wide data frame into a long data frame. Several `key_vars` can be added, however, the new names of the resulting variables should be changed manually. The varying object should be set to NULL, and pererably all variables that should not be turned into th long format should be removed before applying this function.


```{r gather_multiple}

gather_mtpl_df <- gather_multiple(data=data, key_vars=c("id", "id2"), varying = NULL)
head(gather_mtpl_df)

```

### aggregate_df

```{r aggregate_df}

library(padis)
df <- aggregate_df(wide_example_data, id="id")
head(df)

```


## Saving data

```{r write_to_wd, message=FALSE, error=FALSE, eval = FALSE, echo = TRUE}

# If the function is called for the first time and the file does not exist
write_to_wd(data, folder = NULL, type = "xlsx", name="example_data_wfu", overwrite=FALSE)


```


# Section 2 - specific padis functions

Section 2 contains the functions that are specifically built for the tasks at Wake Forest Unviersity and Lara Kammrath's workgroup. They may not be applicable to a broader set of problems. However, they all rely on the functions above.

### match_ir

This function searches for two items that have the same stem but end on two different end-strings, e.g. .r and .i:

```{r match_ir}


library(padis)
data <- imaginer_example_data
head(data)
match_ir(data, var_stem="PSP1.variable")

```

It does not matter if you write `var_stem="PSP1.variable"` or `var_stem="PSP1.variable."` (the difference being the dot at the end), the function works either way. Further, the default for the end-pattern is `".r"` and `".i"`.

### imaginer

This function looks for two items that have the same stem, but different end-strings. It then matches these two variables in the following way:

The first variable with the *.r* at the end, e.g. *PSP1.variable.r*, and the second variable with *.i* at the end, e.g. *PSP1.variable.i*, are scanned and the function identifies which one has a number in it, and if either, put that number in the first new variable, e.g. *PSP1.variable*. A separate 0/1 variable, e.g. *PSP.real.imagined*, would indicate where the numbers in the set of variables came from. 


```{r imaginer}

data <- imaginer_example_data
return_df <- imaginer(data, var_stem="PSP1.variable.", return_complete_data=TRUE)
head(return_df, 10)

```

## Functions for splitting the data frames 

There are two functions to split the raw data frames:

- `get_selection_long`: Turn raw daily data into the selection data set
- `transform_data`: Turn raw daily data and intake data sets into the five specific data sets

First, the data sets need to be loaded (this part is not documented here)

```{r read data, eval = TRUE, echo = FALSE}

## First, the data need to be read:

setwd("/Users/kaihorstmann/Dropbox/01Uni/01Projekte/96collaboration/Lara Kammrath/Materials/data")
data_dy <- read.csv("/Users/kaihorstmann/Dropbox/01Uni/01Projekte/96collaboration/Lara Kammrath/Materials/data/Raw Diary Selection Data.csv", sep = ";", stringsAsFactors = FALSE)
data_in <- xlsx::read.xlsx("/Users/kaihorstmann/Dropbox/01Uni/01Projekte/96collaboration/Lara Kammrath/Materials/data/Raw Intake Sample.xlsx", 1)


```


```{r show intput data}

head(data_dy, 3)
head(data_in, 3)
```


### get_selection_long


```{r get_selection_long}
library(padis)
selection_long <- get_selection_long(data_dy)
head(selection_long, 3)

```


### transform_data


```{r transform data}
# the results will now appear (in the default mode) in the workspace
transform_data(data_dy, data_in, overwrite=TRUE)
```









