---
title: "PADIS vignette"
author: "Kai T. Horstmann"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
---

# Introduction

This package was written for Lara Kammrath from Wake Forest University and her research. It contains some general functions that may apply to multiple, more general problems (section 1) and some very specific functions (section 2) that are tailored to Lara's specific tasks.

It relies on several different packages, which are installed automatically once this package is installed

- tidyr
- data.table
- stringr
- xlsx
- purrr
- psych
- dplyr


## How to install this package:

You can run the following code to install this package from GitHub:


```{r install paids, eval = FALSE, echo = TRUE}


install.packages("devtools")
devtools::install_github("kthorstmann/padis")
library(padis)


```


# Section 1 - General functions

### Example data frame

Th package contains a fake example data frame, `wide_example_data`, which is used to illustrate the functions.

```{r}

library(padis)
data <- wide_example_data
head(data)

```

## Data selection and transformation

### select_vars

This function can be used to select specific variables from a data set. It is primarily build on `grep` or `dplyr::select`.

```{r select_vars}

subset <- select_vars(data, id = "id", prefix="var_s")
head(subset, 3)

subset <- select_vars(data, id = c("id", "id2"), prefix="var_.+3$")
head(subset, 3)

```

To get more information on the use of regular expressions, see `?base::regex`.

### gather_one

```{r gather}

gather_df <- gather_one(data=data, key_vars="id", varying = c("var_p_1", "var_p_2"), new_name="var")
head(gather_df)
# sort by id to make it visible
newdata <- gather_df[order(gather_df$id),] 
head(newdata, 20)


```

### swap_char

This function is used to change variable names since the `gather` function (see below, based on `tidyr::gather`) relies on a specific strucutre of item names. It is primarily used in some functions specific to the problem at hand (section 2).

```{r swap_char}

swap_char("this.string", ".", "_")

```


### gather_multiple

Similar to `gather_one`, `gather_multiple` turns several variables from a wide data frame into a long data frame. Several `key_vars` can be added, however, the new names of the resulting variables should be changed manually. The varying object should be set to NULL, and pererably all variables that should not be turned into th long format should be removed before applying this function.


```{r gather_multiple}

gather_mtpl_df <- gather_multiple(data=data, key_vars=c("id", "id2"), varying = NULL)
head(gather_mtpl_df)

```

### aggregate_df

```{r aggregate_df}

library(padis)
df <- aggregate_df(wide_example_data, id="id")
head(df)

```


Example when doing computations only on a few variables:


```{r aggregate_df example 1}

# Here, we keep only a few variables, i.e. var_p_1 and var_p_2
head(wide_example_data)
df <- aggregate_df(wide_example_data, id="id", 
                   intake_var=c("var_p_1", "var_p_2"))
head(df)

```


Missing data in the ID variable:

```{r aggregate_df with missings}

## we add some missings:

wide_with_missings <- wide_example_data
wide_with_missings[sample(1:nrow(wide_with_missings), 10),"id"] <- NA
wide_with_missings[sample(1:nrow(wide_with_missings), 10),"var_p_1"] <- NA

df <- aggregate_df(wide_with_missings, id="id", 
                   intake_var=c("var_p_1", "var_p_2"))


```


```{r aggregate on special data}
data <- read.csv("/Users/kaihorstmann/Dropbox/01Uni/01Projekte/96collaboration/Lara Kammrath/Materials/data_2/Aggregating.csv")
df <- aggregate_df(data, id="PAR.ID")
head(df)

## return only the sd per participant
df <- aggregate_df(data, id="PAR.ID", out_values="sd")
head(df)

```




## Saving data

```{r write_to_wd, message=FALSE, error=FALSE, eval = FALSE, echo = TRUE}

# If the function is called for the first time and the file does not exist
write_to_wd(data, folder = NULL, type = "xlsx", name="example_data_wfu", overwrite=FALSE)


```


# Section 2 - specific padis functions

Section 2 contains the functions that are specifically built for the tasks at Wake Forest Unviersity and Lara Kammrath's workgroup. They may not be applicable to a broader set of problems. However, they all rely on the functions above.

### match_ir

This function searches for two items that have the same stem but end on two different end-strings, e.g. .r and .i:

```{r match_ir}


library(padis)
data <- imaginer_example_data
data$random <- 2
head(data)
names <- match_ir(data, var_stem = "PSP1.variable")
data[names]


```

It does not matter if you write `var_stem="PSP1.variable"` or `var_stem="PSP1.variable."` (the difference being the dot at the end), the function works either way. Further, the default for the end-pattern is `".r"` and `".i"`.

### imaginer

This function looks for two items that have the same stem, but different end-strings. It then matches these two variables in the following way:

The first variable with the *.r* at the end, e.g. *PSP1.variable.r*, and the second variable with *.i* at the end, e.g. *PSP1.variable.i*, are scanned and the function identifies which one has a number in it, and if either, put that number in the first new variable, e.g. *PSP1.variable*. A separate 0/1 variable, e.g. *PSP.real.imagined*, would indicate where the numbers in the set of variables came from. 


```{r imaginer}

data <- imaginer_example_data
data$PSP1.variable.r[!is.na(data$PSP1.variable.r)] <- 1
data$PSP1.variable.r[is.na(data$PSP1.variable.r)] <- 0
data$PSP1.variable.i[!is.na(data$PSP1.variable.i)] <- 1
data$PSP1.variable.i[is.na(data$PSP1.variable.i)] <- 0

data$PSP1.variable.i + data$PSP1.variable.r

head(data)
return_df <- imaginer(data, var_stem="PSP1.variable.", return_complete_data=TRUE)
head(return_df, 10)

# check if it is zero or 1
# 

```


### merge_multiple_df

This function merges all data up or down into several other data sets. If `merge_down = FALSE`, the data will be merged up, i.e. unique rows of the dataframe `data_from` will be used for the aggregation.

The result is a list with data frames with the same length as `data_list_to`, which has to be a list of data frames as well. The argument `select_variables` allows selecting a specific set of variabels, based on their prefix (this is based on the function `padis::select_vars`).

```{r merge_multiple_df}
library(padis)
## simulate data
data_long_1 <- data.frame(PAR.ID = letters[1:10],
                          var1 = sample(1:10, 10),
                          var2 = sample(1:10, 10))
data_long_2 <- data.frame(PAR.ID = sort(rep(letters[1:10], 3)),
                          var1 = sort(rep(1:10, 3)),
                          var2 = sort(rep(11:20, 3)))
data_long_2_er <- data.frame(PAR.ID = sort(rep(letters[1:10], 3)),
                          var1 = sort(rep(1:10, 3)),
                          var2 = sort(rep(11:20, 3)))
data_long_2_er[3, "var2"] <- 12 ## add a variable that is not the same as the other variables
data_long_3 <- data.frame(PAR.ID = rep(letters[1:10], 4),
                          var1 = sample(1:40, 40),
                          var2 = sample(1:40, 40))
data_short_2 <- data.frame(PAR.ID = rep(letters[1:10], 1),
                          var1 = sample(1:40, 10),
                          var2 = sample(1:40, 10))
data_short_3 <- data.frame(PAR.ID = rep(letters[1:10], 1),
                          var1 = sample(1:40, 10),
                          var2 = sample(1:40, 10))



## merge down, i. e. duplicate rows when going from a higher data set (e.g .level-2) to a lower data set (e.g. level-1), here, also select one additional variable from data_from (this could also be a prefix)
data_list <- merge_multiple_df(data_from = data_long_1, id_var="PAR.ID", list(data_long_2, data_long_3), merge_down = TRUE, select_variables = "var2")
purrr::map(data_list, head)
## merge up, i. e. make the longer data frame short first and merge then
data_list <- merge_multiple_df(data_long_1, id_var="PAR.ID", list(data_short_2, data_short_3), merge_down = FALSE, select_variables = "var2")
purrr::map(data_list, head)

data_list <- merge_multiple_df(data_long_2_er, id_var="PAR.ID", list(data_short_2, data_short_3), merge_down = FALSE, select_variables = "var2")
purrr::map(data_list, head)


```

If variables need to selected first, use the function `select_vars` to select variables from a data frame first, e.g.

```{r example select, eval = FALSE, echo = TRUE}

select_df <- padis::select_vars(data=data, ids="PAR.ID", prefix="the_prefix_of_the_variables")

```


## Functions for splitting the data frames 

There are two functions to split the raw data frames:

- `get_selection_long`: Turn raw daily data into the selection data set
- `transform_data`: Turn raw daily data and intake data sets into the five specific data sets

First, the data sets need to be loaded (this part is not documented here)

```{r read data, eval = TRUE, echo = FALSE}

## First, the data need to be read:

setwd("/Users/kaihorstmann/Dropbox/01Uni/01Projekte/96collaboration/Lara Kammrath/Materials/data")
data_dy <- read.csv("/Users/kaihorstmann/Dropbox/01Uni/01Projekte/96collaboration/Lara Kammrath/Materials/data/Raw Diary Selection Data.csv", sep = ";", stringsAsFactors = FALSE)
data_in <- xlsx::read.xlsx("/Users/kaihorstmann/Dropbox/01Uni/01Projekte/96collaboration/Lara Kammrath/Materials/data/Raw Intake Sample.xlsx", 1)


```


```{r show intput data}

head(data_dy, 3)
head(data_in, 3)
```


### get_selection_long


```{r get_selection_long}
library(padis)
selection_long <- get_selection_long(data_dy)
head(selection_long, 3)

```


### transform_data


```{r transform data}
# the results will now appear (in the default mode) in the workspace
transform_data(data_dy, data_in, overwrite=TRUE)
```



## Example for specific data sets 


```{r }

data <- read.csv2("/Users/kaihorstmann/Dropbox/01Uni/01Projekte/96collaboration/Lara Kammrath/Materials/data_2/KaiTrial.csv", sep = ",")
library(padis)

## this would be for one variable
match_ir(data, var_stem="Issue1.PSP3.ContactChat.")
return_df <- imaginer(data, var_stem="Issue1.PSP3.ContactChat.", return_complete_data=FALSE)
head(return_df)


## if you want to do this now for every variable, do:
names_data <- names(data)## get the names
issue_vars <- grep(x = names_data, pattern = "Issue+.+i$", value = TRUE) ## get the names that end with .i
stems <- stringr::str_replace_all(issue_vars, ".i$", "") ## replace the .i with nothing

list <- list()

for (i in seq_along(stems)) { ## a for loop
  list[[i]] <- imaginer(data, var_stem=stems[i], return_complete_data=FALSE)
}
purrr::map(list, head)

```

### Working with lists

Wotking with lists makes things easier, because multiple data frames can be used at the same time.

```{r examples lists}

## get the head of each data frame in the list object list
purrr::map(data_list, head)

## get the head of each data frame in the list object list
purrr::map(data_list, psych::describe)

## or make your own function
purrr::map(data_list, function(x) {x + 1}) # in this case add one two each variable. If the variable is not numerical, then there is a Warning (and in this case, Missings arise)

```

```{r example day level data, eval = FALSE, echo = FALSE}

ssd.day <- read.csv("/Users/kaihorstmann/Dropbox/01Uni/01Projekte/96collaboration/Lara Kammrath/Materials/data_4/DayLevel 17.08.01.csv")

ssd.agg.day <- aggregate_df(ssd.day, id="PAR.ID", out_values="max")


head(ssd.day)
str(ssd.day)

# ok problem only missing data 
is.na(mean(c(NA, NA, NA), na.rm = TRUE))



```


# match items 

```{r}

ssd.day <- read.csv("/Users/kaihorstmann/Dropbox/01Uni/01Projekte/96collaboration/Lara Kammrath/Materials/data_5/SSD.PracticeWideData.csv")
head(ssd.day)
data <- ssd.day
names <- names(data)
names
library(stringr)
r_variables <- grep(x = names, pattern = "\\.r$", value = TRUE)
i_variables <- grep(x = names, pattern = "\\.i$", value = TRUE)

r_stems <- str_replace(r_variables, "\\.r$", "")
i_stems <- str_replace(i_variables, "\\.i$", "")
stems <- intersect(r_stems, i_stems)

# any(na.omit(data[, "Issue2.PSP16.select.i"]) > 0)

stem <- stems[1]
sum_i_r <- function(stem, data){
  r_var <- paste0(stem, ".r")
  i_var <- paste0(stem, ".i")
  
  if (any(is.na(data[,i_var])) ) {warning(i_var, " contains missing data - please check")}
  if (any(is.na(data[,r_var])) ) {warning(r_var, " contains missing data - please check")}
  out <- rowSums(data[,c(r_var, i_var)], na.rm = FALSE)
  if (any(na.omit(out) > 1 | na.omit(out) < 0)) {warning ("the result contains numbers other than 0 or 1")}
  return(out)
}
sum_i_r(stems[3], data)

library(purrr)

out_sums <- map(stems, ~ sum_i_r(., data))
names(out_sums) <- stems
data.frame(out_sums)
new_data <- cbind(data, out_sums)
head(new_data)

# write.csv(new_data, "/Users/kaihorstmann/Dropbox/01Uni/01Projekte/96collaboration/Lara Kammrath/Materials/data_5/SSD.PracticeWideData_irsummed.csv", row.names=FALSE)

```







